---
title: "Analysis of Longitudinal Pain Questions: Explore"
author: "Matthew J. Kmiecik"
date: last-modified
date-format: "[Last modified:] DD MMM YYYY"
format:
  pdf:
    toc: true
editor: source
editor_options: 
  chunk_output_type: console
execute:
  eval: true
  echo: false
  warning: false
  error: false
  working-directory: project
params:
  version: v1
---

```{r, setup, eval=TRUE}
# libraries ----
library(tidyverse); library(glue); library(kableExtra); library(here)

# functions ----
source("../fns/kable_tools.R") # source("src/fns/kable_tools.R")
#d <- read_rds(here::here("output", "analysis-long-pain-v1.rds"))

# gets filename and loads data
f <- glue("../../output/analysis-long-pain-{params$version}.rds")
d <- read_rds(file = f)
```

## Brief summary

This is version: `r params$version` of this analysis.

This analysis explored the distributions and change over time—using linear mixed models[^1]—of the 4 longitudinal pain questions of interest:

[^1]: I used generalized additive mixed models (GAMs) here. Given that there were no smooth (i.e., non-linear) terms, these became linear mixed models. The advantage of using the GAMs is that the `mgcv` package provides the Tweedie distribution that is useful for zero-inflated data, which is useful for pain ratings that are positively skewed towards zero.

-   `painrating_child_q1` - Over the last 3 months, what was the average cramping pain you experienced with your period (around or below your belly button or lower back)?
-   `painrating_child_q2` - Over the last 2 weeks, how much pelvic pain (NOT including any period pain) have you experienced?
-   `painrating_child_q3` - Please indicate over the last 2 weeks how much pain you have experienced when having a bowel movement.
-   `painrating_child_q4` - Please indicate over the last 2 weeks how much pain you have experienced when emptying your bladder.

Given the results shown below, here were my observations:

-   `painrating_child_q1` shows the most movement over time (increases) and is Gaussian in nature
-   `painrating_child_q2-4` does not seem to change over time and is zero-inflated; therefore special care must be taken when creating a composite variable and using as a dependent variable in linear mixed modeling (or similar) as a Tweedie distribtion fit the data better than a Gaussian distribution.

## Histograms

```{r, fig.height=6}
d$p1
```

## Trajectories

```{r, fig.height=6}
d$p2
```

\pagebreak

## Gaussian linear mixed modeling

```{r}
kable_pdf(data = d$glance_gaus, cap = "Omnibus Model Statistics")
```

```{r}
kable_pdf(d$param_gaus, cap = "Individual Model Estimates", sci_not_cols = "p")
```

```{r, fig.height=5}
d$p4
```

## Tweedie linear mixed modeling

```{r}
kable_pdf(data = d$glance_tw, cap = "Omnibus Model Statistics")
```

```{r}
d$param_tw %>% 
  mutate(est_exp = exp(estimate), .before = se) %>%
  kable_pdf(., cap = "Individual Model Estimates", sci_not_cols = "p")
```

*Note.* Tweedie estimates are on the log scale, so exponentiating the coefficient (`est_exp`) gives the multiplicative change in the expected value of Y for each year increase.

```{r, fig.height=5}
d$p3
```

## Comparing Gaussian vs. Tweedie fit

```{r, fig.height=5}
d$p5
```

