---
title: "Auditory Task Quality Check"
author: "Matthew J. Kmiecik"
date: last-modified
date-format: "[Last modified:] DD MMM YYYY"
format:
  pdf:
    toc: true
    #toc-depth: 3 # Optional: Specifies the number of header levels included
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
  error: false
---

## Brief Summary

This document provides plots of raw and summary data for the auditory task where participants are presented with an auditory stimulus at 5 loudness intensities. These data in general are behaving in the way expected: as the stimuli became louder, participants' ratings of unpleasantness increased. The pattern was observed across all three sessions.

**Note.** These data were filtered to exclude participants with open queries and results may change after resolving those queries (but not by much).

```{r, setup}
# libraries ----
library(tidyverse); library(RColorBrewer)

# data ----
dd <- readRDS(file = "../../output/aud-task-data.rds")
#dd <- readRDS(file = "output/aud-task-data.rds") # uncomment when working in doc

# remove after these data have been cleaned
dc <- dd %>% filter(session %in% c(10:12), rating %in% c(0:20)) 
```

## Histograms

```{r}
# histogram of ratings by session
ggplot(dc, aes(rating)) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous(breaks = seq(0, 20, 5)) +
  labs(x = "Rating", y = "Frequency") +
  facet_wrap(~session) + 
  theme_bw()
```

## Boxplots

```{r}
# boxplot by session
pj <- position_jitter(height = .5, width = .1)
s_cols <- brewer.pal(9, name = "PuBu")[c(6:8)]
names(s_cols) <- c(10:12)
ggplot(
  dc %>% mutate(session = factor(session)), 
  aes(session, rating, color = session, fill = session)
  ) + 
  geom_point(position = pj, alpha = 1/3, shape = 16) +
  geom_boxplot(width = .2, position = position_nudge(x = .3), color = "black") +
  labs(x = "Session", y = "Rating") +
  scale_color_manual(values = s_cols) +
  scale_fill_manual(values = s_cols) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r, fig.height = 5}
# boxplot by stimulus intensity
pj <- position_jitter(height = .5, width = .1)
stim_cols <- brewer.pal(9, name = "BuPu")[c(4:8)]
names(stim_cols) <- c(1:5)
ggplot(
  dc %>% mutate(stim = factor(stim)), 
  aes(stim, rating, group = stim, color = stim, fill = stim)
) + 
  geom_point(position = pj, alpha = 1/3, shape = 16) +
  geom_boxplot(width = .2, position = position_nudge(x = .3), color = "black") +
  labs(x = "Stimulus Intensity", y = "Rating") +
  scale_color_manual(values = stim_cols) +
  scale_fill_manual(values = stim_cols) +
  theme_bw() +
  facet_wrap(~session, ncol = 1) +
  theme(legend.position = "none")
```

## Slopes

```{r}
# subject-wise slopes
pj_num <- .35
pj <- position_jitter(width = pj_num, height = pj_num)
ggplot(
  dc %>% mutate(session = factor(session)), 
  aes(stim, rating, group = ss, color = session)
  ) + 
  geom_point(position = pj, size = .5) +
  geom_line(
    stat = "smooth", method = "lm", se = FALSE, alpha = 1/2, 
    linewidth = .5
    ) +
  theme_bw() +
  labs(x = "Stimulus Intensity", y = " Rating") +
  scale_color_manual(values = s_cols) +
  facet_wrap(~session) +
  theme(legend.position = "none")
```

## Mean Ratings

```{r, fig.height = 5}
# summarises trial-wise
dc_trial_sum <- 
  dc %>% 
  summarise(
    m = mean(rating), sd = sd(rating), n = n(), sem = sd/sqrt(n), 
    .by = c(session, stim)
    )

# plot
pn <- position_nudge(x = .3)
ggplot(
  dc_trial_sum %>% mutate(session = factor(session)), 
  aes(stim, m, group = session, color = session)
  ) +
  geom_point(
    data = dc %>% mutate(session = factor(session)), 
    aes(y = rating), alpha = 1/3, shape = 16, position = pj, size = .75
    ) +
  geom_point(position = pn) +
  geom_errorbar(aes(ymin = m-sem, ymax = m+sem), width = .1, position = pn) +
  geom_line(position = pn) +
  scale_color_manual(values = s_cols) +
  coord_cartesian(ylim = c(0, 20)) +
  labs(
    x = "Stimulus Intensity", y = "Mean Rating", caption = "SEM error bars."
    ) +
  theme_bw() +
  facet_wrap(~session) +
  theme(legend.position = "none")
```

```{r}
library(knitr); library(kableExtra) # libraries for table generation

# preps table data
td <- 
  dc_trial_sum %>% 
  mutate(across(.cols = c(m, sd, sem), .fns = ~round(.x, 2))) %>%
  arrange(session, stim)

# prints table
kable(
  td, format = "latex", booktabs = TRUE, align = "l",
  caption = "Mean Participant Ratings Across Stimulus Intensity and Session"
  ) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )

# inspecting 
# dc %>% filter(session == 12) %>% count(ss) %>% filter(n != 5) %>% pull(ss) -> tss
# dc %>% filter(ss %in% tss, session == 12) %>% View()
```
