---
title: "Visual Acuity Quality Check"
author: "Matthew J. Kmiecik"
date: last-modified
date-format: "[Last modified:] DD MMM YYYY"
format:
  pdf:
    toc: true
    #toc-depth: 3 # Optional: Specifies the number of header levels included
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  warning: false
  error: false
---

## Brief Summary

This document provides plots of raw and summary data for the visual acuity where participants are presented letters to read across 8 trials of decreasing font size. These data in general are behaving in the way expected: as the font size became smaller, the accuracy in reporting the letters decreased. There is a marked decrease in performance after the 6th trial.

This measure may serve as a useful covariate for the visual task.

**Note.** These data were filtered to exclude participants with open queries and results may change after resolving those queries (but not by much).

```{r, setup}
# libraries ----
library(tidyverse); library(RColorBrewer)

# plot elements ----
s_cols <- brewer.pal(9, name = "PuBu")[c(6:8)] # session colors
names(s_cols) <- c(10:12)

# data ----
dd <- readRDS(file = "../../output/visual-acuity-data.rds")
#dd <- readRDS(file = "output/visual-acuity-data.rds") # uncomment when working in doc

# remove after these data have been cleaned
dc <- dd %>% filter(session %in% c(10:12)) 

# summaries ----

# trial summary
dc_trial <- 
  dc %>% 
  summarise(
    p = mean(acc), 
    n = n(), 
    se = sqrt((p * (1 - p)) / n), 
    .by = c(session, trial)
    ) 

# subject-wise summary
dc_ss_sum <- 
  dc %>% 
  summarise(
    p = mean(acc), n = n(), 
    .by = c(session, ss)
    ) 

# session-wise summary
dc_sum <- 
  dc_ss_sum %>%
  summarise(
    P = mean(p), N = n(), SE = sqrt((P * (1 - P)) / N), 
    .by = c(session)
    )
```

## Histograms

```{r}
# histogram of acc by session
ggplot(dc_ss_sum, aes(p)) + 
  geom_histogram(binwidth = .10) +
  scale_x_continuous(breaks = seq(0, 1, .2)) +
  coord_cartesian(xlim = c(0, 1)) +
  labs(x = "Prop. Correct", y = "Frequency") +
  facet_wrap(~session) + 
  theme_bw()
```

## Boxplots

```{r}
# boxplot by session
pj <- position_jitter(height = .025, width = .1)
ggplot(
  dc_ss_sum %>% mutate(session = factor(session)), 
  aes(session, p, color = session, fill = session)
  ) + 
  geom_point(position = pj, alpha = 1/3, shape = 16) +
  geom_boxplot(width = .2, position = position_nudge(x = .3), color = "black") +
  labs(x = "Session", y = "Prop. Correct") +
  scale_color_manual(values = s_cols) +
  scale_fill_manual(values = s_cols) +
  theme_bw() +
  theme(legend.position = "none")
```

# Mean Accuracy

```{r}
# boxplot by trial (i.e., size)
ggplot(
  dc_trial %>% mutate(trial = factor(trial), session = factor(session)), 
  aes(trial, p, group = session, color = session)
) + 
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = p-se, ymax = p+se), width = .1) +
  labs(x = "Trial", y = "Prop. Correct") +
  scale_color_manual(values = s_cols) +
  theme_bw() +
  facet_wrap(~session) +
  theme(legend.position = "none")
```

# Tables

```{r}
library(knitr); library(kableExtra) # libraries for table generation

# preps table data
td <- 
  dc_trial %>% 
  mutate(across(.cols = c(p, se), .fns = ~round(.x, 3))) %>%
  relocate(se, .before = n) %>%
  arrange(session, trial)

# prints table
kable(
  td, format = "latex", booktabs = TRUE, align = "l",
  caption = "Mean Prop. Correct Across Trial (Stimulus Size) and Session"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    full_width = FALSE
  ) %>%
  pack_rows("Baseline", 1, 8) %>%
  pack_rows("PV 1", 9, 16) %>%
  pack_rows("PV 2", 17, 24)

# preps table data
td <- 
  dc_sum %>% 
  mutate(across(.cols = c(P, SE), .fns = ~round(.x, 3))) %>%
  relocate(SE, .before = N) %>%
  arrange(session)

# prints table
kable(
  td, format = "latex", booktabs = TRUE, align = "l",
  caption = "Mean Prop. Correct Across Session"
  ) %>%
  kable_styling(
    latex_options = c("striped", "hold_position"),
    full_width = FALSE
  )
```
